---
title: 'Reproducible Research - Assignment #1'
author: "Andrea Mazzanti"
date: "December 14, 2015"
output: html_document
---

## Loading and preprocessing the data

This is the code I used to read in the data:

```{r, echo = TRUE}
activity <- read.csv("activity.csv")
```
This is other chunk of code used to answer to all the assignment questions:

```{r, echo=TRUE}
daily_steps <- aggregate(activity$steps, by=list(activity$date), FUN=sum, na.rm = TRUE)
colnames(daily_steps) <- c("date", "steps")

daily_intervals <- aggregate(activity$step, by=list(activity$interval), FUN = mean, na.rm = TRUE) 
colnames(daily_intervals) <- c("intervals", "avg_steps")

ind <- which.max(daily_intervals$avg_steps)
int_max <- daily_intervals$intervals[ind]
```

## What is mean total number of steps taken per day?

1. The total number of steps taken per day is:
````{r, echo=TRUE}
print(daily_steps)
````

2. Histogram of the total number of steps taken each day:

````{r, echo=TRUE}
hist(daily_steps$steps, xlab = "Daily steps", main = "Histogram of daily steps", 
     col = "red", breaks = 20)
````

3. Here are the mean and median of the total number of steps taken per day:

````{r, echo=TRUE}
c("mean" = mean(daily_steps$steps, na.rm = TRUE), 
  "median" = median(daily_steps$steps, na.rm = TRUE))
````

## What is the average daily activity pattern?

1. This is the time series plot of the 5-minute interval and the average number of steps taken, averaged across all days:

````{r, echo = TRUE}
plot(x = daily_intervals$intervals, y = daily_intervals$avg_steps, type = "l",
     xlab = "5-minute interval", ylab = "average # of steps")
````


2. The maximum number of steps is obtained at the `r int_max` 5-minute interval.

## Imputing missing values

1.  The total number of missing values in the dataset is:

````{r, echo = TRUE}
NAs <- nrow(activity) - sum(complete.cases(activity))
print(NAs)
````

2. I am going to replace the NAs values in the dataset with the median value for the corresponding 5-minute interval

3. This is the code to create the new dataset, which I called activity_mod:

````{r, echo = TRUE}
## initialize it as the activity dataframe
activity_mod <- activity
## create a dataframe with median per interval
interval_median <- 
        aggregate(activity_mod$steps, by=list(as.factor(activity_mod$interval)), 
                  FUN=median, na.rm = TRUE)
colnames(interval_median) <- c("interval", "median")

## index of NAs
ind2 <- which(is.na(activity_mod$steps))
## find the corresponding index in the interval_median dataframe
ind3 <- match(activity_mod$interval[ind2], interval_median$interval)
## replace NAs
activity_mod$steps[ind2] <-  interval_median$median[ind3]
````

4. This is the histogram of the total number of steps taken each day with the new dataset:

````{r, echo = TRUE}
daily_steps_mod <- aggregate(activity_mod$steps, by=list(activity_mod$date),
                             FUN=sum, na.rm = TRUE)
colnames(daily_steps_mod) <- c("date", "steps")

hist(daily_steps_mod$steps, xlab = "Daily steps", 
     main = "Histogram of daily steps", col = "red", breaks = 20)
````

And these are the mean and the median total number of steps taken per day:

````{r, echo = TRUE}
c("mean" = mean(daily_steps_mod$steps, na.rm = TRUE), 
  "median" = median(daily_steps_mod$steps, na.rm = TRUE))
````

As you can see, these values differ from the first part of the assignment.  
Imputing missing data as I did lowers the mean, but not the median (as expected).

## Are there differences in activity patterns between weekdays and weekends?

1. This is the new factor I created:

````{r, echo = TRUE}
wdays <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
activity_mod$day <- factor(weekdays(as.Date(activity_mod$date, format = "%Y-%m-%d"))
                           %in% wdays,levels=c(FALSE, TRUE), labels=c('weekend', 'weekday'))
````

2. This is the time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days:

````{r, echo = TRUE}
daily_intervals_mod <- aggregate(activity_mod$steps,
                                 by=list(activity_mod$interval,activity_mod$day),
                                 FUN = mean, na.rm = TRUE) 
colnames(daily_intervals_mod) <- c("interval", "day", "mean")

library(lattice)
xyplot(mean ~ interval| day, 
       data = daily_intervals_mod,
       type = "l",
       xlab = "Interval",
       ylab = "Number of steps",
       layout=c(1,2))
````

